Fitness effects of mutations to SARS-CoV-2 proteins

### 2023-08-11 paper/final_submission/paper.tex
made version of `paper.tex` with refs in file and supp excluded for final submission

An important question is whether the mutation fitness estimates are affected by noise from limited statistical sampling of mutations or whether sequencing errors and bioinformatic artefacts distort the estimates.
To assess if this is the case, we repeated the entire fitness estimation using an even larger pre-built UShER mutation-annotated tree of all $\sim$14-million SARS-CoV-2 sequences in GISAID~\cite{shu2017gisaid} as of March-29-2023.
There is an extremely high correlation between fitness estimates made using the $\sim$7-million publicly available sequences and the larger GISAID tree (Figure~\ref{fig:gisaid_corr}).
This concordance indicates that the set of $\sim$7-million public sequences is large enough that doubling the data does not appreciably shift the estimates, and so throughout this paper we use that sequence set due to our preference for publicly available data.
Furthermore, fitness estimates using sequences from specific countries (USA and England, Figure~\ref{fig:corr}B) are also highly concordant, suggesting that sequencing and bioinformatic workflows are not driving the signal.
Lastly, positions known to be under strong constraint {\it a priori} (e.g.~start codons, the ribosomal slippage site) typically have no or only few mutations, suggesting that sequencing errors in consensus sequences are rare.

@@

\begin{figure*}
\includegraphics[width=\linewidth]{figs/dist.pdf}
\caption{
Distribution of effects of different classes of mutations.
{\bf (A)}
Histograms of effects of synonymous, nonsynonymous, and stop-codon mutations across all viral genes.
Neutral mutations have effects of zero (dashed gray vertical lines), and deleterious mutations have negative effects.
{\bf (B)}
Effects of each class of mutation for each viral gene.
Dark squares indicate the median effect, and the lighter rectangles span the interquartile range.
Mutation types are color-coded as in panel A.
The apparent constraint on synonymous mutation in ORF9b is probably because this gene is encoded in an overlapping reading frame with N~\cite{jungreis2021conflicting}.
See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for plots that allow adjustment of the expected-count cutoff and other interactive options (such as separate histograms for each gene).
See Figure~\ref{fig:effects_dist_position_order} for a version of panel B with genes ordered by genomic position rather than constraint on nonsynonymous mutations.
\label{fig:dist}
}
\end{figure*}

The distributions of mutation effects concur with biological intuition about how different classes of mutations impact protein function.
Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
The handful of synonymous or noncoding mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~\ref{fig:synonymous}).

To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}).
SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}; e.g., nsp13 is under stronger protein-level constraint than nsp1).

However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}).
Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).
The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery} and that major variants had stop codons early in ORF8.
Indeed, the loss of accessory proteins such as ORF8 appears to occur with some regularity during the early evolution of non-human viruses in humans~\citep{rochman2022molecular}.
The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~\ref{fig:ORF3a}).
These observations concur with experiments showing SARS-CoV-2 is attenuated by deletion of ORF3a but there is little effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
However, ORF3a's function must be relatively insensitive to its protein sequence, since other than selection against stop codons there is only amino-acid level constraint at a few sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
Observations such as these could help guide experimental studies to better understand protein function.

The accessory proteins are encoded in the last 5kb of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the genes coding for nucleoprotein, E, and M) is under relatively weak purifying selection (Figure~\ref{fig:effects_dist_position_order}).
Note that some accessory proteins also vary in presence/absence across coronaviruses~\citep{v2021coronavirus,llanes2020betacoronavirus}, and are mostly dispensable in cell culture~\citep{mcgrath2022sars,silvas2021contribution,liu2022live,liu2014accessory}.
Decreased purifying selection in genes encoded near the end of the genome is also observed in more distantly related orders of large single-stranded RNA viruses such as closteroviruses~\citep{wang2011genetic, dawson2013virus}.

@@

\subsection*{Mutation-effect estimates better capture functional constraint than dN/dS ratios or predictions from other methods}
A longstanding approach for analyzing protein constraint is to compare rates of nonsynonymous (dN) and synonymous (dS) substitutions at each site~\citep{nielsen1998likelihood,kosakovsky2005not}.
These dN/dS ratios can be calculated by counting mutations or using phylogenetic substitution models~\citep{kosakovsky2005not,yang2000estimating}.
A limitation of dN/dS ratios is they cannot be interpreted in terms of the fitness effects, since they simply represent the relative rate of amino-acid substitution rather than the effects of specific mutations~\citep{spielman2015relationship,kryazhimskiy2008population}.
Nonetheless, we can compare dN/dS ratios to our mutation-effect estimates as measures of the average constraint at each site.
The mutation-effect estimates greatly outperform dN/dS ratios as a measure of site-level constraint as assessed by correlation with deep mutational scanning experiments (Figure~\ref{fig:dnds}).
The reason is in part because some aspects of functional constraint cannot be captured by a dN/dS ratio.
For instance, the ACE2-affinity enhancing spike mutation N501Y arose in several SARS-CoV-2 variants early in the pandemic, and has since remained fixed due its importance for receptor binding~\citep{starr2022deep}.
Our mutation-effect estimates correctly reflect that site 501 is strongly prefers tyrosine, but the site has a high dN/dS ratio due to the early convergent evolution of this site to that preferred amino acid.

Our mutation-effect estimates also correlate better with deep mutational scanning experiments than predictions from two algorithms trained to learn epistatic models of mutation effects from phylogenetically broader but more sparsely sampled sequence data~\cite{thadani2022learning,rodriguez2022epistatic} (Figure~\ref{fig:comparator}).
Our estimates also correlate better with experiments than predictions by a machine-learning algorithm that integrates sequence and epidemiological data~\cite{maher2022predicting} (Figure~\ref{fig:comparator}).
These results suggest that our straightforward approach of directly reading out the effects of mutations from their actual versus expected counts can outperform much more complex models when millions of sequences are available.

\subsection*{Fixed mutations tend to have beneficial or neutral effects}
Amino-acid mutations that have fixed in at least one viral clade are estimated to mostly have neutral or beneficial effects, whereas most other mutations are deleterious (Figure~\ref{fig:fixed_dist}).
This fact is unsurprising: viral lineages that expand into new clades do so because they have acquired beneficial mutations while avoiding deleterious ones~\citep{luksza2014predictive, koelle2015effects, huddleston2020integrating}.
But the fact that the beneficial effects of fixed mutations are correctly estimated by our approach, which simply counts mutation occurrences and does not incorporate information on lineage size, demonstrates such mutations occur independently in many viral lineages that are more successful than average.

Most fixed mutations are estimated to be beneficial regardless of whether estimates are made using all viral clades, or just clades that did not fix the mutation (Figure~\ref{fig:fixed}).
However, a few beneficial fixed mutations show epistatic entrenchment~\citep{shah2015contingency, starr2018pervasive} in the sense that they are not particularly beneficial in clades in which they did not fix (Figure~\ref{fig:fixed}).
The most striking example is S373P in spike, which has experimentally been shown to be neutral or slightly deleterious in pre-Omicron clades, but strongly beneficial in the Omicron clades in which it fixed~\citep{starr2022deep, moulana2022compensatory}.

@@

There are several caveats to our approach.
First, because the number of observations of any given mutation is small compared to the millions of SARS-CoV-2 sequences being analyzed, our approach requires careful quality control to remove sequencing errors.
Second, we assume the rate of each type of nucleotide mutation is uniform across the viral genome, and neglect higher-order context that may influence mutation rate~\citep{sadykov2021short,beale2004comparison}.
Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{huston2021comprehensive,kuo2013functional}---although this probably has only a minor effect, since our analyses show just a handful of synonymous sites are under strong selection (Figure~\ref{fig:synonymous}).
Third, the exact relationship between the statistics we calculate and viral fitness depend on the fraction of all infections that are sequenced (sampling intensity) and viral population dynamics.
Although we derive this relationship, we do not adjust for sampling intensity and population dynamics when estimating mutation effects.
Fourth, we make a single estimate for each mutation across all SARS-CoV-2, neglecting the epistasis that can affect some mutations~\citep{starr2022shifting,moulana2022compensatory}.
Finally there are a few technical caveats to how we count mutations that are discussed in the Methods section.

Conceptually, our approach differs from prior methods that aim to identify beneficial SARS-CoV-2 mutations associated with viral clades that increase in frequency~\citep{obermeyer2022analysis,lee2022inferring,maher2022predicting}.
Those methods draw information primarily from what happens downstream of a mutation.
In contrast, we treat all mutations equivalently regardless of whether they are on a tip node or at the base of a large clade.
Our approach is better for estimating effects of deleterious or nearly neutral mutations, but clade-growth methods may be better for beneficial mutations.
In particular, clade size carries information beyond that contained in mutation counts alone (Figure~\ref{fig:terminal}).
Hopefully future work can combine mutation-counting and clade-growth methods for even better estimates of SARS-CoV-2 mutation effects.
Note our approach is conceptually similar to estimating fitness costs of HIV or polio mutations from mutation-selection balance in deep sequencing of intra-population viral quasispecies~\citep{zanini2017vivo, acevedo2014mutational}, except we analyze mutation occurrences rather than frequencies to account for the phylogenetic structure and genetic hitchhiking that characterize global SARS-CoV-2 evolution.

@@

We then performed quality control by ignoring any branch that met any of the following criteria:
\begin{itemize}
\item it had more than four nucleotide mutations;
\item it contained more than one nucleotide mutation that was a reversion to the Wuhan-Hu-1 reference sequence;
\item it contained more than one nucleotide mutation that was a reversion to the founder sequence for that clade as provided at \url{https://raw.githubusercontent.com/neherlab/SC2_variant_rates/7e738194a8c6592082f1caa9a6ca70cb68289790/data/clade_gts.json} by \cite{neher2022contributions};
\item it contained more than one nucleotide mutation to the same codon.
\end{itemize}
The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors or viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
Because the virus's evolution is very densely sampled, only a small fraction of branches have more than four mutations (Figure~\ref{fig:nmut_counts}).
The rationale for the second and third exclusions is that excess reversions can arise from base-calling pipelines that erroneously call low-coverage sites as reference.
We ignore branches with multiple nucleotide mutations to the same codon (this is very rare) because as detailed below our method is only designed to make estimates for mutations that represent single-nucleotide changes from the clade founder.
Note also that the mutation-annotated tree does not include insertion or deletion mutations, and so we only consider (and make estimates for) point mutations.

@@

\subsection{Comparison to deep mutational scanning}
Deep mutational scanning data were taken from published studies~\citep{dadonaite2023pseudovirus,starr2022deep,flynn2022,flynn2023,iketani2022functional}, using the data at the links specified under the ``dms\_datasets'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}.
For the spike deep mutational scanning~\citep{dadonaite2023pseudovirus} we only included mutations with ``times seen'' values of at least three in the deep mutational scanning.
The RBD data~\citep{starr2022deep} include measurements for two phenotypes (ACE2 affinity and RBD expression), and one of the Mpro studies~\citep{flynn2022} includes measurements for three different phenotypes in yeast (growth, FRET, and transcription factor activity).
Figures~\ref{fig:dms}, \ref{fig:dnds}, and \ref{fig:comparator} shows the effect averaged across all phenotypes measured by each of these studies.
For plots that break the correlations out by phenotype, see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_all_corr.html}.

\subsection{Comparison to dN/dS and other mutation-effect prediction algorithms}
For the comparison to the dN/dS approaches shown in Figure~\ref{fig:dnds}, we used the dN/dS values available at \url{https://github.com/spond/SARS-CoV-2-variation}~\cite{martin2021emergence} for all SARS-CoV-2 sequences, which were calculated using the FEL approach~\cite{kosakovsky2005not}.
Specifically, the file \url{https://raw.githubusercontent.com/spond/SARS-CoV-2-variation/master/windowed-sites-fel-all.csv} contains dN/dS estimates made using large-scale sequence sets from GISAID in three-month windows starting with the earliest sequences from December 2019 and continuing up until Jan-31-2022.
We averaged the dN/dS values for each site over all three month windows, and analyzed those time-window averaged dN/dS values.
In order to ensure comparability in the sequence sets used, for the comparisons in Figure~\ref{fig:dnds}, we used fitness estimates from our approach made using only sequences in the mutation-annotated tree as of Jan-31-2022.
We restricted our analysis to this timeframe because the large-scale dN/dS analyses at \url{https://github.com/spond/SARS-CoV-2-variation} (which use state-of-the-art methods) were only available for that range of dates.

The dN/dS ratios only provide a single number for each site, which cannot be directly compared to either the mutation-effect estimates or the deep mutational scanning, which estimate the effects of individual amino-acid mutations.
We therefore computed site-summary metrics of the mutation-effect estimates and the deep mutational scanning as the average effect of all measured amino-acid mutations at each site, excluding stop codons.
The correlations in Figure~\ref{fig:dnds} are with those site-summary metrics.

We also compared both our mutation-effect estimates and the spike deep mutational scanning measurements~\citep{dadonaite2023pseudovirus} to predictions from three other algorithms:
\begin{itemize}
\item the EpiScores reported by Maher et al (2022)~\cite{maher2022predicting},
\item the DCA mutability scores reported by Rodriguez-Rivas et al (2022)~\cite{rodriguez2022epistatic}, and
\item the EVE scores reported by Thadani et al (2023)~\cite{thadani2022learning}.
\end{itemize}
These comparisons are shown in Figure~\ref{fig:comparator}.
The Maher et al (2022) and Thadani et al (2023) studies report mutation-level predictions and so are compared directly to the deep mutational scanning our mutation-effect estimates; Rodriguez-Rivas et al (2023) report only site-level metrics and so are compared to site-summary metrics as for the dN/dS analysis.

@@

\bibliography{references}
}

\newgeometry{left=1in, right=1in, top=48pt, bottom=60pt}
\onecolumn
\renewcommand{\thepage}{S\arabic{page}}
\setcounter{page}{1}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\clearpage

\section{Supplementary appendix deriving relationship between fitness cost and ratio of expected to actual counts}

With millions of SARS-CoV-2 sequences shared publicly, almost all mutations that are tolerated by the virus are observed dozens to hundreds of times.
Where on the tree and how often on the tree we observe specific mutations has information about the effects of these mutations on viral spread.
The mutation rate depends on the nucleotides involved and possibly on the sequence context and other viral determinants, but for the purpose of this derivation, we will assume the neutral rate $\mu$ is known.
If the mutation is neutral, the total number of times the mutation is observed on the tree is $\mu T$, where $T$ is the total length of the tree (assuming that the mutation never reached high frequency which is true for almost all mutations, particularly when mutations are counted on a per-clade basis relative too the clade founder as done above).

If a mutation reduces fitness, the lineages descending from branches on which this mutation happened will spread more slowly than those without this mutation.
As a result, the down-stream subclades are smaller and more short lived, which in turn means that they will be less likely to be sampled and represented in the tree.
To infer a mutation's effect on fitness, we need to calculate how the probability of observation depends on this fitness effect.

For a mutation to be represented in the tree, one of its descendants has to be sampled and sequenced.
If the total number of descendants is $w$ and the sampling fraction is $\epsilon$, the probability that the mutation is present in the tree is
\begin{equation}
    P = 1 - e^{-w\epsilon}
\end{equation}
$W$ is a random number that depends on the realization of the transmission process, which is commonly modeled by a branching process with birth rate $b$ and death rate $d$.
The death rate here corresponds to clearing an infection, the birth rate to onward transmission. The latter is affected by the fitness cost of the mutation.

>>>>

An important question is whether the mutation fitness estimates are affected by noise from limited statistical sampling of mutations or whether sequencing errors and bioinformatic artefacts distort the estimates.
To assess if this is the case, we repeated the entire fitness estimation using an even larger pre-built UShER mutation-annotated tree of all $\sim$14-million SARS-CoV-2 sequences in GISAID~\cite{shu2017gisaid} as of March-29-2023.
There is an extremely high correlation between fitness estimates made using the $\sim$7-million publicly available sequences and the larger GISAID tree (Figure~S1).
This concordance indicates that the set of $\sim$7-million public sequences is large enough that doubling the data does not appreciably shift the estimates, and so throughout this paper we use that sequence set due to our preference for publicly available data.
Furthermore, fitness estimates using sequences from specific countries (USA and England, Figure~\ref{fig:corr}B) are also highly concordant, suggesting that sequencing and bioinformatic workflows are not driving the signal.
Lastly, positions known to be under strong constraint {\it a priori} (e.g.~start codons, the ribosomal slippage site) typically have no or only few mutations, suggesting that sequencing errors in consensus sequences are rare.

@@

\begin{figure*}
\includegraphics[width=\linewidth]{figs/dist.pdf}
\caption{
Distribution of effects of different classes of mutations.
{\bf (A)}
Histograms of effects of synonymous, nonsynonymous, and stop-codon mutations across all viral genes.
Neutral mutations have effects of zero (dashed gray vertical lines), and deleterious mutations have negative effects.
{\bf (B)}
Effects of each class of mutation for each viral gene.
Dark squares indicate the median effect, and the lighter rectangles span the interquartile range.
Mutation types are color-coded as in panel A.
The apparent constraint on synonymous mutation in ORF9b is probably because this gene is encoded in an overlapping reading frame with N~\cite{jungreis2021conflicting}.
See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for plots that allow adjustment of the expected-count cutoff and other interactive options (such as separate histograms for each gene).
See Figure~S3 for a version of panel B with genes ordered by genomic position rather than constraint on nonsynonymous mutations.
\label{fig:dist}
}
\end{figure*}

The distributions of mutation effects concur with biological intuition about how different classes of mutations impact protein function.
Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
The handful of synonymous or noncoding mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~S2).

To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figures~\ref{fig:dist}B and S3).
SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figures~\ref{fig:dist}B and S3; e.g., nsp13 is under stronger protein-level constraint than nsp1).

However, most accessory proteins are under little constraint (Figures~\ref{fig:dist}B and S3).
Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).
The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery} and that major variants had stop codons early in ORF8.
Indeed, the loss of accessory proteins such as ORF8 appears to occur with some regularity during the early evolution of non-human viruses in humans~\citep{rochman2022molecular}.
The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~S4).
These observations concur with experiments showing SARS-CoV-2 is attenuated by deletion of ORF3a but there is little effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
However, ORF3a's function must be relatively insensitive to its protein sequence, since other than selection against stop codons there is only amino-acid level constraint at a few sites like 135 and 138 (Figure~S4).
Observations such as these could help guide experimental studies to better understand protein function.

The accessory proteins are encoded in the last 5kb of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the genes coding for nucleoprotein, E, and M) is under relatively weak purifying selection (Figure~S3).
Note that some accessory proteins also vary in presence/absence across coronaviruses~\citep{v2021coronavirus,llanes2020betacoronavirus}, and are mostly dispensable in cell culture~\citep{mcgrath2022sars,silvas2021contribution,liu2022live,liu2014accessory}.
Decreased purifying selection in genes encoded near the end of the genome is also observed in more distantly related orders of large single-stranded RNA viruses such as closteroviruses~\citep{wang2011genetic, dawson2013virus}.

@@

\subsection*{Mutation-effect estimates better capture functional constraint than dN/dS ratios or predictions from other methods}
A longstanding approach for analyzing protein constraint is to compare rates of nonsynonymous (dN) and synonymous (dS) substitutions at each site~\citep{nielsen1998likelihood,kosakovsky2005not}.
These dN/dS ratios can be calculated by counting mutations or using phylogenetic substitution models~\citep{kosakovsky2005not,yang2000estimating}.
A limitation of dN/dS ratios is they cannot be interpreted in terms of the fitness effects, since they simply represent the relative rate of amino-acid substitution rather than the effects of specific mutations~\citep{spielman2015relationship,kryazhimskiy2008population}.
Nonetheless, we can compare dN/dS ratios to our mutation-effect estimates as measures of the average constraint at each site.
The mutation-effect estimates greatly outperform dN/dS ratios as a measure of site-level constraint as assessed by correlation with deep mutational scanning experiments (Figure~S5).
The reason is in part because some aspects of functional constraint cannot be captured by a dN/dS ratio.
For instance, the ACE2-affinity enhancing spike mutation N501Y arose in several SARS-CoV-2 variants early in the pandemic, and has since remained fixed due its importance for receptor binding~\citep{starr2022deep}.
Our mutation-effect estimates correctly reflect that site 501 is strongly prefers tyrosine, but the site has a high dN/dS ratio due to the early convergent evolution of this site to that preferred amino acid.

Our mutation-effect estimates also correlate better with deep mutational scanning experiments than predictions from two algorithms trained to learn epistatic models of mutation effects from phylogenetically broader but more sparsely sampled sequence data~\cite{thadani2022learning,rodriguez2022epistatic} (Figure~S6).
Our estimates also correlate better with experiments than predictions by a machine-learning algorithm that integrates sequence and epidemiological data~\cite{maher2022predicting} (Figure~S6).
These results suggest that our straightforward approach of directly reading out the effects of mutations from their actual versus expected counts can outperform much more complex models when millions of sequences are available.

\subsection*{Fixed mutations tend to have beneficial or neutral effects}
Amino-acid mutations that have fixed in at least one viral clade are estimated to mostly have neutral or beneficial effects, whereas most other mutations are deleterious (Figure~S7).
This fact is unsurprising: viral lineages that expand into new clades do so because they have acquired beneficial mutations while avoiding deleterious ones~\citep{luksza2014predictive, koelle2015effects, huddleston2020integrating}.
But the fact that the beneficial effects of fixed mutations are correctly estimated by our approach, which simply counts mutation occurrences and does not incorporate information on lineage size, demonstrates such mutations occur independently in many viral lineages that are more successful than average.

Most fixed mutations are estimated to be beneficial regardless of whether estimates are made using all viral clades, or just clades that did not fix the mutation (Figure~S8).
However, a few beneficial fixed mutations show epistatic entrenchment~\citep{shah2015contingency, starr2018pervasive} in the sense that they are not particularly beneficial in clades in which they did not fix (Figure~S8).
The most striking example is S373P in spike, which has experimentally been shown to be neutral or slightly deleterious in pre-Omicron clades, but strongly beneficial in the Omicron clades in which it fixed~\citep{starr2022deep, moulana2022compensatory}.

@@

There are several caveats to our approach.
First, because the number of observations of any given mutation is small compared to the millions of SARS-CoV-2 sequences being analyzed, our approach requires careful quality control to remove sequencing errors.
Second, we assume the rate of each type of nucleotide mutation is uniform across the viral genome, and neglect higher-order context that may influence mutation rate~\citep{sadykov2021short,beale2004comparison}.
Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{huston2021comprehensive,kuo2013functional}---although this probably has only a minor effect, since our analyses show just a handful of synonymous sites are under strong selection (Figure~S2).
Third, the exact relationship between the statistics we calculate and viral fitness depend on the fraction of all infections that are sequenced (sampling intensity) and viral population dynamics.
Although we derive this relationship, we do not adjust for sampling intensity and population dynamics when estimating mutation effects.
Fourth, we make a single estimate for each mutation across all SARS-CoV-2, neglecting the epistasis that can affect some mutations~\citep{starr2022shifting,moulana2022compensatory}.
Finally there are a few technical caveats to how we count mutations that are discussed in the Methods section.

Conceptually, our approach differs from prior methods that aim to identify beneficial SARS-CoV-2 mutations associated with viral clades that increase in frequency~\citep{obermeyer2022analysis,lee2022inferring,maher2022predicting}.
Those methods draw information primarily from what happens downstream of a mutation.
In contrast, we treat all mutations equivalently regardless of whether they are on a tip node or at the base of a large clade.
Our approach is better for estimating effects of deleterious or nearly neutral mutations, but clade-growth methods may be better for beneficial mutations.
In particular, clade size carries information beyond that contained in mutation counts alone (Figure~S9).
Hopefully future work can combine mutation-counting and clade-growth methods for even better estimates of SARS-CoV-2 mutation effects.
Note our approach is conceptually similar to estimating fitness costs of HIV or polio mutations from mutation-selection balance in deep sequencing of intra-population viral quasispecies~\citep{zanini2017vivo, acevedo2014mutational}, except we analyze mutation occurrences rather than frequencies to account for the phylogenetic structure and genetic hitchhiking that characterize global SARS-CoV-2 evolution.

@@

We then performed quality control by ignoring any branch that met any of the following criteria:
\begin{itemize}
\item it had more than four nucleotide mutations;
\item it contained more than one nucleotide mutation that was a reversion to the Wuhan-Hu-1 reference sequence;
\item it contained more than one nucleotide mutation that was a reversion to the founder sequence for that clade as provided at \url{https://raw.githubusercontent.com/neherlab/SC2_variant_rates/7e738194a8c6592082f1caa9a6ca70cb68289790/data/clade_gts.json} by \cite{neher2022contributions};
\item it contained more than one nucleotide mutation to the same codon.
\end{itemize}
The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors or viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
Because the virus's evolution is very densely sampled, only a small fraction of branches have more than four mutations (Figure~S10).
The rationale for the second and third exclusions is that excess reversions can arise from base-calling pipelines that erroneously call low-coverage sites as reference.
We ignore branches with multiple nucleotide mutations to the same codon (this is very rare) because as detailed below our method is only designed to make estimates for mutations that represent single-nucleotide changes from the clade founder.
Note also that the mutation-annotated tree does not include insertion or deletion mutations, and so we only consider (and make estimates for) point mutations.

@@

\subsection{Comparison to deep mutational scanning}
Deep mutational scanning data were taken from published studies~\citep{dadonaite2023pseudovirus,starr2022deep,flynn2022,flynn2023,iketani2022functional}, using the data at the links specified under the ``dms\_datasets'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}.
For the spike deep mutational scanning~\citep{dadonaite2023pseudovirus} we only included mutations with ``times seen'' values of at least three in the deep mutational scanning.
The RBD data~\citep{starr2022deep} include measurements for two phenotypes (ACE2 affinity and RBD expression), and one of the Mpro studies~\citep{flynn2022} includes measurements for three different phenotypes in yeast (growth, FRET, and transcription factor activity).
Figures~\ref{fig:dms}, S5, and S6 shows the effect averaged across all phenotypes measured by each of these studies.
For plots that break the correlations out by phenotype, see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_all_corr.html}.

\subsection{Comparison to dN/dS and other mutation-effect prediction algorithms}
For the comparison to the dN/dS approaches shown in Figure~S5, we used the dN/dS values available at \url{https://github.com/spond/SARS-CoV-2-variation}~\cite{martin2021emergence} for all SARS-CoV-2 sequences, which were calculated using the FEL approach~\cite{kosakovsky2005not}.
Specifically, the file \url{https://raw.githubusercontent.com/spond/SARS-CoV-2-variation/master/windowed-sites-fel-all.csv} contains dN/dS estimates made using large-scale sequence sets from GISAID in three-month windows starting with the earliest sequences from December 2019 and continuing up until Jan-31-2022.
We averaged the dN/dS values for each site over all three month windows, and analyzed those time-window averaged dN/dS values.
In order to ensure comparability in the sequence sets used, for the comparisons in Figure~S5, we used fitness estimates from our approach made using only sequences in the mutation-annotated tree as of Jan-31-2022.
We restricted our analysis to this timeframe because the large-scale dN/dS analyses at \url{https://github.com/spond/SARS-CoV-2-variation} (which use state-of-the-art methods) were only available for that range of dates.

The dN/dS ratios only provide a single number for each site, which cannot be directly compared to either the mutation-effect estimates or the deep mutational scanning, which estimate the effects of individual amino-acid mutations.
We therefore computed site-summary metrics of the mutation-effect estimates and the deep mutational scanning as the average effect of all measured amino-acid mutations at each site, excluding stop codons.
The correlations in Figure~S5 are with those site-summary metrics.

We also compared both our mutation-effect estimates and the spike deep mutational scanning measurements~\citep{dadonaite2023pseudovirus} to predictions from three other algorithms:
\begin{itemize}
\item the EpiScores reported by Maher et al (2022)~\cite{maher2022predicting},
\item the DCA mutability scores reported by Rodriguez-Rivas et al (2022)~\cite{rodriguez2022epistatic}, and
\item the EVE scores reported by Thadani et al (2023)~\cite{thadani2022learning}.
\end{itemize}
These comparisons are shown in Figure~S6.
The Maher et al (2022) and Thadani et al (2023) studies report mutation-level predictions and so are compared directly to the deep mutational scanning our mutation-effect estimates; Rodriguez-Rivas et al (2023) report only site-level metrics and so are compared to site-summary metrics as for the dN/dS analysis.

@@

\begin{thebibliography}{76}

\bibitem[{Abdool~Karim {\rm and} de~Oliveira}(2021)]{abdool2021new}
{Abdool~Karim, S.~S. {\rm and} T.~de~Oliveira}, 2021 New {SARS-CoV-2}
  variants—clinical, public health, and vaccine implications. New England
  Journal of Medicine {\bf 384}: 1866--1868.

\bibitem[{Acevedo {\em et~al.\/}}(2014){Acevedo, Brodsky, {\rm and}
  Andino}]{acevedo2014mutational}
{Acevedo, A., L.~Brodsky, {\rm and} R.~Andino}, 2014 Mutational and fitness
  landscapes of an {RNA} virus revealed through population sequencing. Nature
  {\bf 505}: 686--690.

\bibitem[{Aksamentov {\em et~al.\/}}(2021){Aksamentov, Roemer, Hodcroft, {\rm
  and} Neher}]{aksamentov2021nextclade}
{Aksamentov, I., C.~Roemer, E.~B. Hodcroft, {\rm and} R.~A. Neher}, 2021
  Nextclade: clade assignment, mutation calling and quality control for viral
  genomes. Journal of Open Source Software {\bf 6}: 3773.

\bibitem[{Beale {\em et~al.\/}}(2004){Beale, Petersen-Mahrt, Watt, Harris,
  Rada, {\rm and} Neuberger}]{beale2004comparison}
{Beale, R.~C., S.~K. Petersen-Mahrt, I.~N. Watt, R.~S. Harris, C.~Rada, {\em
  et~al.\/}}, 2004 {Comparison of the differential context-dependence of DNA
  deamination by APOBEC enzymes: correlation with mutation spectra in vivo}.
  Journal of Molecular Biology {\bf 337}: 585--596.

\bibitem[{Bhatt {\em et~al.\/}}(2021){Bhatt, Scaiola, Loughran, Leibundgut,
  Kratzel, Meurs, Dreos, O’Connor, McMillan, Bode, {\em
  et~al.\/}}]{bhatt2021structural}
{Bhatt, P.~R., A.~Scaiola, G.~Loughran, M.~Leibundgut, A.~Kratzel, {\em
  et~al.\/}}, 2021 Structural basis of ribosomal frameshifting during
  translation of the {SARS-CoV-2 RNA} genome. Science {\bf 372}: 1306--1313.

\bibitem[{Bloom {\em et~al.\/}}(2023){Bloom, Beichman, Neher, {\rm and}
  Harris}]{bloom2023evolution}
{Bloom, J.~D., A.~C. Beichman, R.~A. Neher, {\rm and} K.~Harris}, 2023
  Evolution of the {SARS-CoV-2} mutational spectrum. Molecular Biology and
  Evolution {\bf 40}: msad085.

\bibitem[{Cao {\em et~al.\/}}(2022{a}){Cao, Jian, Wang, Yu, Song, Yisimayi, An,
  Chen, Zhang, Wang, {\em et~al.\/}}]{cao2022imprinted}
{Cao, Y., F.~Jian, J.~Wang, Y.~Yu, W.~Song, {\em et~al.\/}}, 2022{a} {Imprinted
  SARS-CoV-2 humoral immunity induces convergent Omicron RBD evolution}. Nature
  {\bf DOI 10.1038/s41586-022-05644-7}.
